---
title: Playing with a new R package; welcome ggcharts!
author: Carlos Vecina Tebar
date: "2020-04-01"
slug: ggcharts-lollipop-plot-in-r
translationKey: "ggcharts-lollipop-plot-in-r"
output:
  blogdown::html_page:
    highlight: tango
categories:
- R
- Tips
tags: []
subtitle: 'In this post we download and try out the new CRAN package ggcharts. A ggplot extension with some default oneliner plots.'
summary: "Thomas Neitman's ggcharts"
authors: [carlos-vecina]
featured: false
draft: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
```


Yesterday I was playing with the new and first of many Thomas Neitman's CRAN package!

Its documentation reminded me that more than a year ago we did a talk series about text mining. In one of them, we analized the reviews of the most important spanish banks' APPs.

One of the plots was a lollipop chart showing the users score in the Play Store, and its deviation from the mean. I used to remember this plot cooler than it actually seems to me lol
It's maybe telling us that we have improved our skills. Hope it.

```{r}
library(tidyverse)
library(purrr)

# Data Prep
reviews_score <- fst::read_fst("reviews_score.fst")
globalMean <- round(mean(reviews_score$Mean),2)
center_formatter <- function(x) {x + globalMean}   # Set the mean as a center
custom_labs <- labs(title="Ranking Spanish Banking APPs reviews (1 to 5 stars) ", 
                    subtitle=paste0("Filtering out reviews without comment. Last week data. The mean score is: ", globalMean), 
                    x="",
                    y="Score")


ggplot(reviews_score, aes(x=Bank, y=MeanDev, label=Mean)) + 
  geom_segment(aes(x = Bank, y = 0,
                   yend = MeanDev, xend = Bank,
                   col=AboveBelow),size=2)+ 
  geom_point(stat='identity', aes(col=AboveBelow), size = 14)  +
  guides(colour = guide_legend(override.aes = list(size=2))) +
  scale_color_manual(name="Scoring", 
                     labels = c("Above"="Above the mean",
                                "Below"="Below the mean"), 
                     values = c("Above"="#00ba38",
                                "Below"="#f8766d")) + 
  geom_text(color="black", size=5) +
  coord_flip() + 
  theme_minimal() +
  scale_y_continuous(labels=center_formatter) +
  custom_labs

```

Well, it's quite good if you don't pay attention to the big circles and to the long code to write such a "standart" chart.

Here is where **ggcharts** shines. And it's only the 0.1.0 version. But having reusable plot templates but in a flexible way seems such a cool idea for me, 

See an three-liner aprox to the above plot:
```{r}
# install.packages("ggcharts") :)
library(ggcharts)

  lollipop_chart(reviews_score, x=Bank, y=MeanDev, color=AboveBelow) +
  scale_color_manual(name="Scoring", 
                     labels = c("Above"="Above the mean","Below"="Below the mean"), 
                     values = c("Above"="#00ba38", "Below"="#f8766d")) +
  scale_y_continuous(labels=center_formatter) + custom_labs 
```

That was so cool and quite much cleaner.

There are several paths to keep improving its capabilities. So here are some enancements. Tell him yours... **see you in his ggcharts [repo.](https://github.com/thomas-neitmann/ggcharts)!**

<br>

Proposed enhancements:

- Easy & Quick auto-aligned Annotations:
```{r}
reviews_score %>% 
  lollipop_chart(x=Bank, y=MeanDev, limit = 25, #line_color = colours,
                 highlight = "BBVA",point_color = "blue") + 
  scale_y_continuous(labels=center_formatter) +
  annotate("segment", x = 5, xend = 5.3, y = 0.55, yend = 0.7, colour = "black") +
  annotate("text", x = 5.4, y = 0.75, label = "4.05") + custom_labs
```

- Easy Image axis and plot insertion:
```{r}
# https://gist.github.com/jonocarroll/2f9490f1f5e7c82ef8b791a4b91fc9ca
# https://stackoverflow.com/questions/54973129/including-images-on-axis-label-in-an-animated-ggplot2
```

- Easy non 0 centered in lolliplot and diverging lolliplot:
```{r, eval=FALSE}
# center parameter?
scale_y_continuous(labels = function(x) {x + center})
```

- Color aes and highlight conflict
```{r, eval=FALSE}
# Counter intuitive line_color parameter when highlight
colours <- c("green, green", "green", "red", "red", "red")
lollipop_chart(reviews_score, x=Bank, y=MeanDev, limit = 25, line_color = colours) # BBVA is green
lollipop_chart(reviews_score, x=Bank, y=MeanDev, limit = 25, line_color = colours,
                 highlight = "BBVA") # the highlighting BBVA now red

diverging_lollipop_chart(reviews_score, Bank, Mean) # not working with all positive values (linked with  to center in non 0 value)
```

<br>

Enjoy ggcharts!

<br>
